---
title: "Introduction to spatio-temporal modelling for environment and ecology"
subtitle: "VAST theory"
date: "March 2024"
fontsize: 10pt
bibliography : ["ref.bib"]
csl : ices-journal-of-marine-science.csl
output:
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "default"
    slide_level: 3
    # template: mytemplate.tex
    includes:
      in_header: header-simple.tex
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(mapdata)
library(sf)

# Map
mapBase <- map("worldHires", fill = T, plot = F)
mapBase <- st_as_sf(mapBase) %>% filter(ID %in% c("France","Spain","UK","Ireland"))
grid_projection <- "+proj=longlat +datum=WGS84"

```

# Introduction

\frametitle{Some illustrations}

\center

\includegraphics[width=1\textwidth]{images/simu1.png}

------------------------------------------------------------------------

\center

\includegraphics[width=1\textwidth]{images/simu2.png}

------------------------------------------------------------------------

```{r,include=F,fig.align='center'}

# EVHOE data
load("data/EVHOE_2008_2019.RData")

# Haul data
Haul_df <- Save_Datras$datras_HH.full %>%
  dplyr::select(Year,long,lati,StNo,HaulNo,Depth)

# Extent of the EVHOE domain
xlims <- range(pretty(Haul_df$long))
ylims <- range(pretty(Haul_df$lati))

# Catch data
Catch_df <- Save_Datras$datras_sp.HL.full %>%
  group_by(Year,long,lati,StNo,HaulNo,scientificname) %>%
  dplyr::summarise(CatchWgt = CatCatchWgt,
                   TotNum = TotalNo)

# Join with haul data to add missing hauls to catch data
Catch_df_2 <- full_join(Catch_df,Haul_df) %>%
  filter(scientificname == "Merluccius_merluccius")
Catch_df_2$CatchWgt[which(is.na(Catch_df_2$CatchWgt))] <- 0

# Plot
Evhoe_plot <- ggplot(Catch_df_2)+
  geom_point(aes(x=long,y=lati,col=CatchWgt))+
  scale_color_distiller(palette="Spectral",trans="log10")+
  facet_wrap(.~Year)+
  geom_sf(data=mapBase)+
  coord_sf(xlim = xlims, ylim = ylims, expand = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5,face = "bold",size=14),
        panel.spacing.x = unit(4, "mm"))+
  ggtitle("Merluccius merluccius (EVHOE)",subtitle=" ")+
  ylab("")+xlab("")

```

```{r,warning=F,echo=F,fig.align='center',out.width="70%"}

plot(Evhoe_plot)

```

------------------------------------------------------------------------

\center

\includegraphics[width=0.6\linewidth]{"images/co2.png"}

------------------------------------------------------------------------

\frametitle{Data and ecological processes}

\vspace{\baselineskip}

What are the characteristics of these data?

\medskip

```{=tex}
\onslide<2-4>{

\begin{itemize}

\item They arise from the ecological process of interest (= they are \textbf{conditionnal} on the ecological process)

\item \textbf{Noisy} (= they are not perfect observations of the ecological process)

\item \textbf{Sparse} (= they do not cover the full area and some hypothesis need to be set to predict the ecological process between the sampled locations)

\end{itemize}

}
```
\vspace{\baselineskip}

What are the characteristics of the ecological process we want to infer?

```{=tex}
\onslide<3-4>{

\begin{itemize}

\item \textbf{Hidden} or \textbf{latent}

\item \textbf{Structured} (relations that structure the process?)

\end{itemize}

}
```
\vspace{\baselineskip}

How to relate the data to the ecological process?

\vspace{\baselineskip}

```{=tex}
\onslide<4>{

\begin{center}
\ding{224} \textbf{Hierarchical modeling}
\end{center}

}
```

------------------------------------------------------------------------

\frametitle{Hierarchical models}

```{=tex}
\begin{columns}
\begin{column}{0.4\textwidth}

Let's define observations $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}}$:
$$\textcolor{BaptisteLightGreen}{\boldsymbol{Y}} | \textcolor{BaptisteBlue}{\boldsymbol{S}}, \textcolor{red}{\boldsymbol{\theta}} \sim \mathcal{L}_Y(\textcolor{BaptisteBlue}{\boldsymbol{S}},\textcolor{red}{\boldsymbol{\theta}_{obs}})$$

a latent field $\textcolor{BaptisteBlue}{\boldsymbol{S}}$:
$$\textcolor{BaptisteBlue}{\boldsymbol{S}} | \textcolor{red}{\boldsymbol{\theta}_{process}} \sim \mathcal{L}_S(\textcolor{red}{\boldsymbol{\theta}_{process}})$$

and parameters:
$$\textcolor{red}{\boldsymbol{\theta}} = (\textcolor{red}{\boldsymbol{\theta}_{obs}},\textcolor{red}{\boldsymbol{\theta}_{process}})$$

\vspace{\baselineskip}

\scriptsize

There can be several data sources $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}_1}$ and $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}_2}$, in which case each has its own probability distribution ($\mathcal{L}_{Y_1}$,$\mathcal{L}_{Y_2}$) and observation parameters ($\textcolor{red}{\boldsymbol{\theta}_{obs , 1}}$, $\textcolor{red}{\boldsymbol{\theta}_{obs , 2}}$).

\end{column}
\begin{column}{0.6\textwidth}

\center{\includegraphics[width=0.9\columnwidth]{images/dag_ssm.png}}

\end{column}
\end{columns}
```

------------------------------------------------------------------------

```{=tex}
\begin{center}
\ding{224} \textbf{How to specify such model for space-time applications?}
\end{center}
```
\medskip

\tableofcontents

\medskip

```{=tex}
\pause
\scriptsize
```
Based on the package \href{https://github.com/James-Thorson-NOAA/VAST}{VAST} (Vector Autoregressive Spatio-Temporal) coded by Jim Torson (NOAA, Seattle)

# Univariate spatio-temporal models

\scriptsize

Let's denote:

\medskip

```{=tex}
\begin{columns}
\begin{column}{0.5\textwidth}

\begin{itemize}
\item $\boldsymbol{Y} = (Y_i, i \in \{1, \cdots , n\})$ the observation vector

\item $\boldsymbol{S} = (S_i, i \in \{1, \cdots , n\})$ the latent field vector

\item $\boldsymbol{\delta} = (\delta_i, i \in \{1, \cdots , n\})$ the latent variables
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item $\boldsymbol{\beta} = (\beta_j, j \in \{1, \cdots , p\})$ the parameter vector

\item $\boldsymbol{X} _ {( n \times p )}$ the covariate matrix

\item $f$ a link function.
\end{itemize}
\end{column}
\end{columns}
```
\normalsize

\bigskip \bigskip

$$Y | \boldsymbol{\delta}, \boldsymbol{\beta} \sim \mathcal{L}(\boldsymbol{S}, \sigma^2)$$

$$f(\boldsymbol{S}) = \boldsymbol{X} \cdot \boldsymbol{\beta} + \boldsymbol{\delta}$$

$$\boldsymbol{\delta} \sim \mathcal{MG}(0,\boldsymbol{\Sigma})$$

\bigskip

\center

The key element is the covariance matrix $\boldsymbol{\Sigma}_{(n \times n)}$

------------------------------------------------------------------------

\large

$$\boldsymbol{\Sigma} = \begin{pmatrix}
Var[\delta_1] & Cov[\delta_1,\delta_2] & \cdots & Cov[\delta_1,\delta_n] \\ 
Cov[\delta_2,\delta_1] & Var[\delta_2] & \ddots & \vdots \\
\vdots & \ddots & \ddots & \vdots \\
Cov[\delta_n,\delta_1] & \cdots & \cdots & Var[\delta_n] \\
\end{pmatrix}$$

\center

\medskip

\includegraphics[width=0.7\linewidth]{"images/domain.png"}

------------------------------------------------------------------------

\center

\ding{224} \textbf{The covariance is stationary (and isotropic)} \begin{align}
Cov[\delta_i,\delta_j] = Cov[h_{ij}] \text{ with } h_{ij} \in \mathbb{R} \nonumber \\
h_{ij} \text{ is the distance between i and j} \nonumber
\end{align}

\includegraphics[width=0.7\linewidth]{"images/altern_cov.png"}

\medskip

\small

We will denote $\delta_i$ as $\delta(x_i)$ in the following

------------------------------------------------------------------------

```{=tex}
\begin{center}

\includegraphics[width=0.7\textwidth]{images/lindgren1.png}

\includegraphics[width=0.7\textwidth]{images/lindgren2.png}

\end{center}
```
\tiny

Lindgren, F., Rue, H., and Lindström, J. (2011). An explicit link between Gaussian fields and Gaussian Markov random fields: the stochastic partial differential equation approach. Journal of the Royal Statistical Society Series B: Statistical Methodology, 73(4), 423-498.

------------------------------------------------------------------------

\frametitle{Moving to spatio-temporal}

The ideas are similar, but we add temporal correlations in the expression of the random effect $\boldsymbol{\delta}$. \medskip

```{=tex}
\begin{center}
\includegraphics[width=0.9\linewidth]{"images/st_maps.png"}
\end{center}
```
\vspace{\baselineskip}

\scriptsize

Let's introduce the model:

$$\delta(x,t)=\varphi \cdot \delta(x,t-1) + \omega(x,t) \text{ for } t = 2,...,T$$

-   $\varphi \in ]-1;1[$ is the autoregressive temporal term

-   $\omega(x,t)$ is a purely spatial GRF

-   $\omega(x,1)$ derives from the stationary distribution $\mathcal{N}(0,\sigma^2 / (1 - \varphi ^ 2))$

------------------------------------------------------------------------

```{=tex}
\begin{columns}
\begin{column}{0.65\textwidth}

\begin{center}

\includegraphics[width=0.9\textwidth]{images/seabass.png}

\scriptsize

Average monthly distribution of sea bass in the Bay of Biscay from 2008 to 2018

\end{center}

\end{column}
\begin{column}{0.35\textwidth}

\scriptsize

Alglave Baptiste, Olmos et al. (\textit{under review}). Investigating fish adult phenology through Empirical Orthogonal Functions

\end{column}
\end{columns}
```

------------------------------------------------------------------------

```{=tex}
\begin{columns}
\begin{column}{0.5\textwidth}

\includegraphics[width=1\textwidth]{images/thorson1.png}

\end{column}
\begin{column}{0.5\textwidth}

\scriptsize

Thorson, J. T., Shelton, A. O., Ward, E. J., and Skaug, H. J. (2015). Geostatistical delta-generalized linear mixed models improve precision for estimated abundance indices for West Coast groundfishes. ICES Journal of Marine Science, 72(5), 1297-1310.

\end{column}
\end{columns}
```
# Multivariate spatio-temporal models

\tableofcontents[currentsection]

------------------------------------------------------------------------

\frametitle{Multivariate modelling}

Often, spatial data are multivariate

\textit{e.g.} in survey data, several species are recorded in a single haul.

\quad \textbf{\ding{224} How to extract the common patterns between these species?}

```{r,include=F,fig.align='center'}

Catch_df_3 <- full_join(Catch_df,Haul_df) %>%
  filter(scientificname %in% c("Merluccius_merluccius","Lepidorhombus_whiffiagonis","Micromesistius_poutassou","Trachurus_trachurus","Lophius_piscatorius","Lophius_budegassa"))

# Plot
Evhoe_multi_plot <- ggplot(Catch_df_3)+
  geom_point(aes(x=long,y=lati,col=TotNum))+
  scale_color_distiller(palette="Spectral",trans="log10")+
  facet_wrap(.~scientificname)+
  geom_sf(data=mapBase)+
  coord_sf(xlim = xlims, ylim = ylims, expand = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5,face = "bold",size=14),
        panel.spacing.x = unit(4, "mm"))+
  ggtitle("Some important species of the EVHOE survey")+
  ylab("")+xlab("")

```

```{r,warning=F,echo=F,fig.align='center',out.width="65%"}

plot(Evhoe_multi_plot)

```

------------------------------------------------------------------------

\frametitle{How to model jointly these species?}

\large

Let's now denote $\boldsymbol{S}_{(n \times l)}$ the matrix of observations of species in $n$ locations for $l$ species.

$$\log(\boldsymbol{S}_{i,\cdot})=\boldsymbol{X} \cdot \boldsymbol{\beta} + \boldsymbol{L} \boldsymbol{\Delta}^T_{i,\cdot}$$

\bigskip

$\rightarrow \log(\boldsymbol{S}_{i,\cdot})$ is the log-expected densities for all species $l$ at locations $i \in \{1,\cdots, n\}$. \bigskip

$\rightarrow \boldsymbol{X} \cdot \boldsymbol{\beta}$ is the trend part. \bigskip

$\rightarrow \boldsymbol{L} \boldsymbol{\Delta}^T_{i,\cdot}$ is the part of the model that controls species (or variables) interaction.

\bigskip

\center

\ding{224} How is it parameterized?

------------------------------------------------------------------------

\frametitle{How to model jointly these species?}

$\boldsymbol{\Delta}_{(l \times k)}$ is the matrix containing the spatial factors. $\boldsymbol{L}_{(l,k)}$ contains the loading factors. \bigskip

\underline{\textbf{Latent variables ($\boldsymbol{\Delta}$):}} \medskip

$\boldsymbol{\Delta}^T_{i,\cdot}$ is the value of $\boldsymbol{\Delta}$ for all spatial factor $k$ at location $i$. $\Delta$ is a matrix where each column represents one of the spatial factor. Each latent variable $\boldsymbol{\delta}_j$ is parameterized as a GRF. \medskip

$$\boldsymbol{\Delta} = (\boldsymbol{\delta}_1,\boldsymbol{\delta}_2,\cdots,\boldsymbol{\delta}_k) \text{   where    } \boldsymbol{\delta}_j \sim \mathcal{MG}(0,\boldsymbol\Sigma)$$

\bigskip

\underline{\textbf{Loading factors ($\boldsymbol{L}$):}} \medskip

The loadings are contained in $\boldsymbol{L}_{(l,k)}$. It represents the linear relationship between the latent variables in $\boldsymbol{\Delta}$ and the log-expected intensity $\log(\boldsymbol{S})$. \medskip

\scriptsize

$$
\mathbf{L}=\left[\begin{array}{cccc}
l_{1,1} & 0 & \ldots & 0 \\
l_{2,1} & l_{2,2} & \ldots & 0 \\
l_{3,1} & l_{3,2} & \ldots & 0 \\
\ldots & \ldots & \ldots & \ldots \\
l_{l, 1} & l_{l, 2} & \ldots & l_{l, k}
\end{array}\right]
$$

------------------------------------------------------------------------

\includegraphics[width=0.925\textwidth]{images/thorson345.png}

\medskip

\tiny

Thorson, J. T., Scheuerell, M. D., Shelton, A. O., See, K. E., Skaug, H. J., & Kristensen, K. (2015). Spatial factor analysis: a new tool for estimating joint species distributions and correlations in species range. Methods in Ecology and Evolution, 6(6), 627-637.

------------------------------------------------------------------------

\frametitle{How to model jointly these species?}

<!-- \large -->

The interpretation of $\Delta$ is complicated because $\boldsymbol{L}$ is triangular inferior to ensure identifiability (something pretty common in dynamic factor analysis). \bigskip

Interpretation can be simplified by rotating the loading factors and the spatial factors.

\bigskip

$$
\begin{aligned}
& \boldsymbol{\Delta}^{\prime}=\boldsymbol{H} \Delta \\
& \boldsymbol{L}^{\prime}=\boldsymbol{L H}^{-1}
\end{aligned}
$$

\bigskip

$H$ can be computed in several ways.

-   through standard diagonalization ($\equiv$ EOF/PCA)

-   Varimax rotation $\rightarrow$ pushing the spatial factors towards $0$ or $\pm 1$.

-   Quartimin rotation $\rightarrow$ same but drop constraints of orthogonality.

\medskip

The last two are used to minimize the number of factors needed to explain $\boldsymbol{\Delta}$

------------------------------------------------------------------------

\frametitle{How to model jointly these species?}

The last point is the covariance among species / variables.

$$Cov[\Delta_{(\cdot,j)},\Delta_{(\cdot,j')}] = \sum^k_{p=m} l_{j m} l_{j^{\prime} m}$$

$$Cor[\Delta_{(\cdot,j)},\Delta_{(\cdot,j')}] = \frac{\sum^k_{m=1} l_{j m} l_{j^{\prime} m}}{\sqrt{\sum^k_{m=1} l_{j m}^2} \sqrt{\sum^k_{m=1} l_{j^{\prime} m}^2}}$$

\medskip

\center

\includegraphics[width=0.35\textwidth]{images/thorson6.png}

# Model inference

\tableofcontents[currentsection]

------------------------------------------------------------------------

\frametitle{How to infer such model?}

\large

Standard methods are not efficient.

\bigskip

The keystone is the likelihood:

$$L_M(\textcolor{red}{\boldsymbol{\theta}}) = P (\textcolor{BaptisteGreen}{\boldsymbol{Y}} | \textcolor{red}{\boldsymbol{\theta}}) = \int_{\mathbb{R}^{q}} P (\textcolor{BaptisteGreen}{\boldsymbol{Y}}, \textcolor{BaptisteBlue}{\boldsymbol{\delta}} | \textcolor{red}{\boldsymbol{\theta}}) d\textcolor{BaptisteBlue}{\boldsymbol{\delta}}$$

\medskip

\footnotesize

$\textcolor{BaptisteGreen}{\boldsymbol{Y}}$ are the observations, $\textcolor{red}{\boldsymbol{\theta}}$ are the parameters, $\textcolor{BaptisteBlue}{\boldsymbol{\delta}}$ are the latent random variables.

\normalsize

\medskip

\pause

In a spatio-temporal context $q$ can be very high \smallskip

\scriptsize

\quad  \quad \quad \ding{224} require efficient numerical methods (1) to bypass the integration step,

\quad \quad \quad \phantom{\ding{224}} (2) to reduce the dimensionality of $\textcolor{BaptisteBlue}{\boldsymbol{\delta}}$ and (3) to derive efficiently the likelihood.

\normalsize

\medskip

\pause

Three methods to enhance computing:

-   Laplace approximation (= approximation of the likelihood)

-   SPDE approach (= approximation of the Gaussian random effect)

-   Automatic differentiation (= efficient derivation technics)

\medskip

\footnotesize

\pause

\textbf{\underline{R package:}} `R-INLA` for bayesian inference (Rue et al., 2017) and `TMB` for maximum likelihood inference (Kristensen et al., 2015).

------------------------------------------------------------------------

```{=tex}
\begin{columns}
\begin{column}{0.5\textwidth}

\begin{center}
\textbf{SPDE approach}
\end{center}

\end{column}
\begin{column}{0.5\textwidth}

\begin{center}
\textbf{Laplace approximation}
\end{center}

\end{column}
\end{columns}
```
```{=tex}
\begin{columns}
\begin{column}{0.5\textwidth}

\center

\scriptsize 

\textbf{\underline{First set of results}}

\tiny

\medskip

A Gaussian field wih Matérn Covariance can be represented as a Gauss-Markov field \ding{224} sparse representation of the spatial effect

\begin{center}
\includegraphics[width=0.35\textwidth]{images/spde.png}
\end{center}

\scriptsize 

\textbf{\underline{Second set of results}}

\tiny

\includegraphics[width=0.65\textwidth]{images/spde1.png}
\includegraphics[width=0.65\textwidth]{images/spde2.png}


\end{column}
\begin{column}{0.5\textwidth}

\scriptsize \medskip

\textbf{\underline{Conditions required:}}

\begin{itemize}
\item normality of the random effect $\delta$
\item some regularity conditions on $f$ (\textit{e.g.} only one maximum)
\end{itemize}

\begin{center}

\includegraphics[width=1\textwidth]{images/gaussian_laplace.png} \\
\scriptsize
\textbf{The true marginal (\textcolor{BaptisteBlue}{blue}), the Laplace approximation (\textcolor{red}{red}) and the Gaussian approximation (\textcolor{BaptisteOrange}{orange}).}

\end{center}

\end{column}
\end{columns}
```
# Towards more complex methods

\tableofcontents[currentsection]

------------------------------------------------------------------------

\frametitle{Towards more complex methods}

\medskip

-   Modelling population dynamics \medskip

-   Modelling movement \medskip

In both cases, this becomes a state space model where:

$$S(t,s)=g\left(S(t-1,s)\right) \cdot e^{\varepsilon(t,s)}$$

where: \medskip

\quad ${\varepsilon}(t,s) \sim \mathcal{MN}(0,\boldsymbol{\Sigma})$, \medskip

\quad $g:\mathbb{R}^L \rightarrow \mathbb{R}^L$ the function modeling population dynamics or movement. \medskip

\begin{center}
\includegraphics[width=0.4\textwidth]{images/state_space.png}
\end{center}


------------------------------------------------------------------------

\frametitle{Modelling population dynamics}

\center

\only<1>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model1.png}}

\only<2>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model2.png}}

\only<3>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model3.png}}

\only<4>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model4.png}}

\only<5>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model5.png}}

\only<6>{\includegraphics[width=0.85\textwidth]{images/hierarchical_model6.png}}

------------------------------------------------------------------------

\center

\includegraphics[width=1\textwidth]{images/spatial_dyn.png}

------------------------------------------------------------------------

\frametitle{Modelling movement}

\large

Movement will be modelled by a matrix $\mathbf{M}^*_{(n\times n)}$ that is instantaneous movement rate. $\mathbf{M}^*$ is used to model movement from one time step $t$ to another. \bigskip

$\mathbf{M}$ is a combination of 3 matrices:

$$\mathbf{M}^* = \mathbf{D}^* + \mathbf{Z}^* + \mathbf{V}^*$$

\medskip

where $\mathbf{D}^*$ is \underline{diffusion}, $\mathbf{Z}^*$ is \underline{preference towards a specific habitat}, $\mathbf{V}^*$ is \underline{advection}. \medskip

\bigskip

Then to get the movement matrix $\mathbf{M}(t) = \exp (\mathbf{M}^* \cdot \Delta t)$, where $\Delta t$ is the duration of the interval
starting at time $t$. \medskip

All columns of $\mathbf{M}(t)$ sum to 1.

------------------------------------------------------------------------

\begin{center}
\includegraphics[width=0.85\textwidth]{images/movement.png}
\end{center}

\tiny

Thorson, J. T., Barbeaux, S. J., Goethel, D. R., Kearney, K. A., Laman, E. A., Nielsen, J. K., ... & Thompson, G. G. (2021). Estimating fine‐scale movement rates and habitat preferences using multiple data sources. Fish and Fisheries, 22(6), 1359-1376.

------------------------------------------------------------------------

\frametitle{Take home message}

\large

\textbf{Main ideas:} \medskip

-   \underline{hierarchical} modelling (not geostatistics per se)

\medskip

-   spatial/temporal \underline{correlation}

\medskip

-   multivariate analysis \medskip

\quad \ding{224} extract \underline{structuring/shared patterns}, and their variation in time \medskip

\quad \ding{224} quantify \underline{correlation among species/variables}

\bigskip

\center

\textbf{Many applications in ecology (and other fields)} \medskip

\textbf{Many statistical developments too} \medskip

\textbf{But may require new inference methods...} \medskip


------------------------------------------------------------------------


\Large

\center

\textbf{Appendices}

------------------------------------------------------------------------

\center

\includegraphics[width=0.7\linewidth]{"images/covariogram.png"}
