---
title: "Introduction to spatio-temporal modelling for environment and ecology"
subtitle: "VAST theory"
date: "March 2024"
fontsize: 10pt
bibliography : ["ref.bib"]
csl : ices-journal-of-marine-science.csl
output:
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "default"
    slide_level: 3
    # template: mytemplate.tex
    includes:
      in_header: header-simple.tex
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(corrplot)
library(dplyr)
library(fields)
library(geoR)
library(ggplot2)
library(gstlearn)
library(INLA)
library(mapdata)
library(maps)
library(paletteer)
library(plot.matrix)
library(sf)
library(sp)
library(splancs)
library(tidyr)

run_simu <- F

# Map
mapBase <- map("worldHires", fill = T, plot = F)
mapBase <- st_as_sf(mapBase) %>% filter(ID %in% c("France","Spain","UK","Ireland"))
grid_projection <- "+proj=longlat +datum=WGS84"

```


# Introduction

\frametitle{Some illustrations}

\center
\includegraphics[width=1\textwidth]{images/simu1.png}

---

\center
\includegraphics[width=1\textwidth]{images/simu2.png}

---


```{r,include=F,fig.align='center'}

# EVHOE data
load("data/EVHOE_2008_2019.RData")

# Haul data
Haul_df <- Save_Datras$datras_HH.full %>%
  dplyr::select(Year,long,lati,StNo,HaulNo,Depth)

# Extent of the EVHOE domain
xlims <- range(pretty(Haul_df$long))
ylims <- range(pretty(Haul_df$lati))

# Catch data
Catch_df <- Save_Datras$datras_sp.HL.full %>%
  group_by(Year,long,lati,StNo,HaulNo,scientificname) %>%
  dplyr::summarise(CatchWgt = CatCatchWgt)

# Join with haul data to add missing hauls to catch data
Catch_df_2 <- full_join(Catch_df,Haul_df) %>%
  filter(scientificname == "Merluccius_merluccius")
Catch_df_2$CatchWgt[which(is.na(Catch_df_2$CatchWgt))] <- 0

# Plot
Evhoe_plot <- ggplot(Catch_df_2)+
  geom_point(aes(x=long,y=lati,col=CatchWgt))+
  scale_color_distiller(palette="Spectral",trans="log10")+
  facet_wrap(.~Year)+
  geom_sf(data=mapBase)+
  coord_sf(xlim = xlims, ylim = ylims, expand = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(hjust = 0.5,face = "bold",size=14),
        panel.spacing.x = unit(4, "mm"))+
  ggtitle("Merluccius merluccius (EVHOE)",subtitle=" ")+
  ylab("")+xlab("")

```


```{r,warning=F,echo=F,fig.align='center',out.width="70%"}

plot(Evhoe_plot)

```

---

\center

\includegraphics[width=0.6\linewidth]{"images/co2.png"}

---

\frametitle{Data and ecological processes}


\vspace{\baselineskip}

What are the characteristics of these data?

\medskip

\onslide<2-4>{

\begin{itemize}

\item They arise from the ecological process of interest (= they are \textbf{conditionnal} on the ecological process)

\item \textbf{Noisy} (= they are not perfect observations of the ecological process)

\item \textbf{Sparse} (= they do not cover the full area and some hypothesis need to be set to predict the ecological process between the sampled locations)

\end{itemize}

}

\vspace{\baselineskip}

What are the characteristics of the ecological process we want to infer?

\onslide<3-4>{

\begin{itemize}

\item \textbf{Hidden} or \textbf{latent}

\item \textbf{Structured} (relations that structure the process?)

\end{itemize}

}

\vspace{\baselineskip}

How to relate the data to the ecological process?

\vspace{\baselineskip}

\onslide<4>{

\begin{center}
\ding{224} \textbf{Hierarchical modeling}
\end{center}

}

---

\frametitle{Hierarchical models}


```{=tex}
\begin{columns}
\begin{column}{0.4\textwidth}

Let's define observations $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}}$:
$$\textcolor{BaptisteLightGreen}{\boldsymbol{Y}} | \textcolor{BaptisteBlue}{\boldsymbol{S}}, \textcolor{red}{\boldsymbol{\theta}} \sim \mathcal{L}_Y(\textcolor{BaptisteBlue}{\boldsymbol{S}},\textcolor{red}{\boldsymbol{\theta}_{obs}})$$

a latent field $\textcolor{BaptisteBlue}{\boldsymbol{S}}$:
$$\textcolor{BaptisteBlue}{\boldsymbol{S}} | \textcolor{red}{\boldsymbol{\theta}_{process}} \sim \mathcal{L}_S(\textcolor{red}{\boldsymbol{\theta}_{process}})$$

and parameters:
$$\textcolor{red}{\boldsymbol{\theta}} = (\textcolor{red}{\boldsymbol{\theta}_{obs}},\textcolor{red}{\boldsymbol{\theta}_{process}})$$

\vspace{\baselineskip}

\scriptsize

There can be several data sources $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}_1}$ and $\textcolor{BaptisteLightGreen}{\boldsymbol{Y}_2}$, in which case each has its own probability distribution ($\mathcal{L}_{Y_1}$,$\mathcal{L}_{Y_2}$) and observation parameters ($\textcolor{red}{\boldsymbol{\theta}_{obs , 1}}$, $\textcolor{red}{\boldsymbol{\theta}_{obs , 2}}$).

\end{column}
\begin{column}{0.6\textwidth}

\center{\includegraphics[width=0.9\columnwidth]{images/dag_ssm.png}}

\end{column}
\end{columns}
```

---

\begin{center}
\ding{224} \textbf{How to specify such model for space-time applications?}
\end{center}

\medskip

\tableofcontents

\medskip

\pause
\scriptsize

Based on the package \href{https://github.com/James-Thorson-NOAA/VAST}{VAST} (Vector Autoregressive Spatio-Temporal) coded by Jim Torson (NOAA, Seattle)

# Univariate spatio-temporal models

\scriptsize

Let's denote:

\medskip

\begin{columns}
\begin{column}{0.5\textwidth}

\begin{itemize}
\item $\boldsymbol{Y} = (Y_i, i \in \{1, \cdots , n\})$ the observation vector

\item $\boldsymbol{S} = (S_i, i \in \{1, \cdots , n\})$ the latent field vector

\item $\boldsymbol{\delta} = (\delta_i, i \in \{1, \cdots , n\})$ the latent variables
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item $\boldsymbol{\beta} = (\beta_j, j \in \{1, \cdots , p\})$ the parameter vector

\item $\boldsymbol{X} _ {( n \times p )}$ the covariate matrix

\item $f$ a link function.
\end{itemize}
\end{column}
\end{columns}

\normalsize

\bigskip \bigskip

$$Y | \boldsymbol{\delta}, \boldsymbol{\beta} \sim \mathcal{L}(\boldsymbol{S}, \sigma^2)$$

$$f(\boldsymbol{S}) = \boldsymbol{X} \cdot \boldsymbol{\beta} + \boldsymbol{\delta}$$

$$\boldsymbol{\delta} \sim \mathcal{MG}(0,\boldsymbol{\Sigma})$$

\bigskip 

\center
The key element is the covariance matrix $\boldsymbol{\Sigma}_{(n \times n)}$

---

\large

$$\boldsymbol{\Sigma} = \begin{pmatrix}
Var[\delta_1] & Cov[\delta_1,\delta_2] & \cdots & Cov[\delta_1,\delta_n] \\ 
Cov[\delta_2,\delta_1] & Var[\delta_2] & \ddots & \vdots \\
\vdots & \ddots & \ddots & \vdots \\
Cov[\delta_n,\delta_1] & \cdots & \cdots & Var[\delta_n] \\
\end{pmatrix}$$

\center

\medskip

\includegraphics[width=0.7\linewidth]{"images/domain.png"}

---

\center
\ding{224} \textbf{The covariance is stationary (and isotropic)}
\begin{align}
Cov[\delta_i,\delta_j] = Cov[h_{ij}] \text{ with } h_{ij} \in \mathbb{R} \nonumber \\
h_{ij} \text{ is the distance between i and j} \nonumber
\end{align}


\includegraphics[width=0.7\linewidth]{"images/altern_cov.png"}

---

\center
\includegraphics[width=0.7\linewidth]{"images/covariogram.png"}

---

\frametitle{Moving to spatio-temporal}

The ideas are similar, but we add temporal correlations in the expression of the random effect $\boldsymbol{\delta}$. \medskip

\begin{center}
\includegraphics[width=0.9\linewidth]{"images/st_maps.png"}
\end{center}

\vspace{\baselineskip}

\scriptsize

Let's introduce the model:

$$\delta(x,t)=\varphi \cdot \delta(x,t-1) + \omega(x,t) \text{ for } t = 2,...,T$$

- $\varphi \in ]-1;1[$ is the autoregressive temporal term

- $\omega(x,t)$ is a purely spatial GRF

- $\omega(x,1)$ derives from the stationary distribution $\mathcal{N}(0,\sigma^2 / (1 - \varphi ^ 2))$

---

\frametitle{How to infer such model?}

\large

Standard methods are not efficient.

\bigskip

The keystone is the likelihood:

$$L_M(\textcolor{red}{\theta}) = P (\textcolor{BaptisteGreen}{Y} | \textcolor{red}{\theta}) = \int_{\mathbb{R}^{q}} P (\textcolor{BaptisteGreen}{Y}, \textcolor{BaptisteBlue}{\delta} | \textcolor{red}{\theta}) d\textcolor{BaptisteBlue}{\delta}$$


\medskip

\footnotesize

$\textcolor{BaptisteGreen}{Y}$ are the observations, $\textcolor{red}{\theta}$ are the parameters, $\textcolor{BaptisteBlue}{\delta}$ are the latent variables.

\normalsize

\medskip

\pause

In a spatio-temporal context $q$ can be very high \smallskip

\scriptsize 
\quad  \quad \quad \ding{224} require efficient numerical methods (1) to bypass the integration step, 

\quad \quad \quad \phantom{\ding{224}} (2) to reduce the dimensionality of $\textcolor{BaptisteBlue}{\delta}$ and (3) to derive efficiently the likelihood. 

\normalsize

\medskip

\pause

Three methods to enhance computing:

- Laplace approximation (= approximation of the likelihood)

- SPDE approach (= approximation of the Gaussian random effect)

- Automatic differentiation (= efficient derivation technics)

\medskip

\footnotesize

\pause

\textbf{\underline{R package:}} `R-INLA` for bayesian inference (Rue et al., 2017) and `TMB` for maximum likelihood inference (Kristensen et al., 2015).

---

\begin{columns}
\begin{column}{0.5\textwidth}

\begin{center}
\textbf{SPDE approach}
\end{center}

\end{column}
\begin{column}{0.5\textwidth}

\begin{center}
\textbf{Laplace approximation}
\end{center}

\end{column}
\end{columns}




\begin{columns}
\begin{column}{0.5\textwidth}

\center

\scriptsize 

\textbf{\underline{First set of results}}

\tiny

\medskip

A Gaussian field wih Matérn Covariance can be represented as a Gauss-Markov field \ding{224} sparse representation of the spatial effect

\begin{center}
\includegraphics[width=0.35\textwidth]{images/spde.png}
\end{center}

\scriptsize 

\bigskip

\textbf{\underline{Second set of results}}

\tiny

\only<1>{\includegraphics[width=0.75\textwidth]{images/spde1.png}}
\only<2>{\includegraphics[width=0.75\textwidth]{images/spde2.png}}


\end{column}
\begin{column}{0.5\textwidth}

\scriptsize \medskip

\textbf{\underline{Conditions required:}}

\begin{itemize}
\item normality of the random effect $\delta$
\item some regularity conditions on $f$ (\textit{e.g.} only one maximum)
\end{itemize}

\begin{center}

\includegraphics[width=1\textwidth]{images/gaussian_laplace.png} \\
\scriptsize
\textbf{The true marginal (\textcolor{BaptisteBlue}{blue}), the Laplace approximation (\textcolor{red}{red}) and the Gaussian approximation (\textcolor{BaptisteOrange}{orange}).}

\end{center}

\end{column}
\end{columns}


# Multivariate spatio-temporal models

test

# Dimension reduction for spatio-temporal models

test

# Model inference

test

# Conclusion

test
